Gene expression profiling of reproductive meristem types in early rice inflorescences by laser microdissection

### 2016-02-03 ms.Rmd
typo
Comments stefan helene james

---
title: Gene expression profiling of reproductive meristem identities in rice inflorescence
csl: csl/the-plant-journal.csl
header-includes:
  \usepackage{graphicx}
  \usepackage[utf8]{inputenc}
  \DeclareUnicodeCharacter{2265}{$\geq$}
output:
  word_document:
    fig_caption: yes
    reference_docx: csl/wordStyles.docx
  BiocStyle::html_document:
    dev: CairoPDF
    fig_caption: yes
    force_captions: yes
    highlight: pygments
    number_sections: yes
bibliography: bib/papersLib.bibtex
---

@@

In rice, the start of the reproductive phase and subsequent initiation of inflorescence development involves fast morphological transformations. After the establishment of the rachis meristem (RM), some cells differentiate into primary branch meristem (PBM) for the formation and elongation of primary branches. During primary branch elongation, the PBM can give rise to axillary meristems (AM), which may differentiate into secondary and higher-order branches or be converted directly to spikelet meristem (SM). The PBM and secondary branch meristem (SBM) are both converted to a terminal SM. Each terminal or axillary SM produces one floret meristem (FM), which differentiates into a single floret [@Ikeda:2004tk]. A detailed histological analysis of rice inflorescences was performed to select four morphologically distinct inflorescence meristem types for laser microdissection (LMD). After differentiation of the flag leaf, the first stage of reproductive development is the conversion of the shoot apical meristem (SAM) to the RM. The first bract primordium was produced opposite the flag leaf (`r I(pasteLabel("Figure ", figCount, "f_lmdFigure"))`a). After the first stage of reproductive growth the rachis meristem produced PBMs in the axils of newly developed bracts (`r I(pasteLabel("Figure ", figCount, "f_lmdFigure"))`b). After the formation of primary branches, RM activity was lost, bract growth ceased and primary branches elongated (ePBM) (`r I(pasteLabel("Figure ", figCount, "f_lmdFigure"))`c). Each PBM produced a number of AMs as they elongated. For this analysis, SM differentiation was considered as the final stage of panicle development (`r I(pasteLabel("Figure ", figCount, "f_lmdFigure"))`d). The RM, PBM and ePBM/AM stages are the indeterminate stages in which meristematic cells are maintained, whilst the SM has a determinate fate in which the stem cell activity will be lost and from which florets will differentiate [@Ikeda:2004tk].

@@

Several previous transcriptomic studies have measured gene expression in whole rice inflorescences at later stages of development than those described here [*e.g.* @Wang:2010im; @Sato:2011eh; @Sharma:2012kq; @Khanday:2013bz; @Jiang:2014im]. Another study used microarray experiments at earlier stages to highlight the importance of transcription factors in panicle development [@Furutani:2006hx]. LMD has also been used for precise control of developmental stage during the collection of whole panicle sections for microarrays [@Kobayashi:2012fi]. This identified three MADS genes, which are members of the *SQUA*-like (*FUL*-like) clade, that are co-expressed with *PANICLE PHYTOMER2* (*PAP2*), and a quadruple knockdown of these four genes resulted in defects in inflorescence development [@Kobayashi:2012fi]. All four genes were detected in all stages in the RNAseq datset (Data S1), confirming their expression in reproductive meristems. In contrast to previous transcriptomic studies, only the meristematic regions of young panicles were collected by LMD for the RNAseq dataset presented in this article, to restrict the measurement of gene expression to those tissues.

@@

Clusters `r getClusterName('dec', 'small')` and `r getClusterName('inc', 'small')` contain genes that change in expression between apical (RM) and axillary meristems (PBM to SM). Axillary meristem initiation requires auxin synthesis and transport in *Arabidopsis thaliana* and maize [reviewed by @Gallavotti:2013ki], and excess auxin may be inactivated by conjugation to amino acids by GH3 enzymes [@Staswick:2005cv], but these processes are not as well understood in rice. Cluster `r getClusterName('inc', 'small')` contains the auxin response factor *ARF6A* and the small auxin-up RNA *SAUR33*, as well as two ABC transporters (*ALS1* and *MDR12*), while the complementary cluster `r getClusterName('dec', 'small')` contains four uncharacterized genes with annotations relating to auxin (*IAA10* and *IAA16*, two auxin-responsive transcriptional regulators; *ABCB27*, an ABC transporter; and *LOC_Os01g63770*, an AUX/LAX transporter), and the GH3 gene related to auxin homeostasis, *GH3-8*. Transgenic rice plants presumably expressing *GH3-8* in the endogenous pattern of *FLO-LFY HOMOLOG OF RICE* (*RFL*) [*i.e.* PBM and SBM; @IkedaKawakatsu:2012co] have smaller panicles with fewer branches [@Yadav:2011db], and GH3-8 conjugates amino acids to auxin *in vitro* [@Ding:2008hv], suggesting that the panicle phenotype may be related to reduced auxin signalling. Although auxin response factors and Aux/IAA response genes are generally more strongly expressed at earlier stages (`r I(pasteLabel("Figure ", figCount, "f_gsea"))`), the presence of auxin-related genes in complementary clusters suggests that the transcriptional effects of auxin signalling are dependent on meristem type. Understanding of the relationship between auxin and meristem specification will require functional analysis of the components of the regulatory network involved in the response in these tissues. Nevertheless, the difference in expression of annotated transcription factor genes between RM and PBM samples and the recovery of co-expression clusters of primarily uncharacterized genes that appear to be either switched on or off between the RM and the PBM indicate a significant change in gene expression between apical and axillary meristems.

Two *STMADS11*-like (*SVP*-like) MADS genes, *MADS22* and *MADS55*, decrease in expression between PBM and SM, whilst the third (*MADS47*) was not detected (`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`a). *MADS22* and *MADS55* are activated by the *G1*-Like (G1L/ALOG) transcriptional activator *G1L5* (*TAWAWA1*/*TAW1*), but activation of *MADS47* has not been reported [@Yoshida:2013ff]. Overexpression of *TAW1* increases the production of secondary branches and leads to the formation of tertiary branches, resulting in higher grain yield [@Yoshida:2013ff]. *TAW1* was recovered from the same cluster as *MADS55*, and the presence and co-expression of genes that are involved in inflorescence development along with other TF genes indicates that the clusters capture biologically meaningful groups of genes, including uncharacterized signalling components. Only three ALOG genes (which are annotated with the Pfam domain PF04852) were detected in the dataset. Two were retrieved from clusters `r getClusterName('dec', 'small')` and `r getClusterName('dec', 'dec')` (*G1L2* and *TAW1* respectively), and the third (*G1L1*) has a similar expression pattern (`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`b). *taw1* missense mutants have small inflorescences with reduced grain yield, but this phenotype was stronger when *TAW1* was targeted by RNA interference [@Yoshida:2013ff]. This may suggest functional redundancy between the three co-expressed ALOG genes, *G1L1*, *G1L2* and *TAW1*, with more than one affected by the RNAi construct used against *TAW1*. Although the increase in secondary branches linked to *TAW1* overexpression suggests that it acts by suppression of SM identity, perhaps via downstream factors that may promote BM activity such as the *STMADS11*-like genes, *taw1* plants also have fewer primary branches [@Yoshida:2013ff], and the peak expression in the RM of *G1L1*, *G1L2*, *TAW1*, *MADS22* and *MADS55* suggests an additional role before the transition from apical to axillary meristem.

@@

Compared to the switch from apical to axillary meristem, more gradual changes in gene expression were apparent during the sequential transition from PBM to SM (`r I(pasteLabel("Figure ", figCount, "f_mfuzzClusters"))`). This transition corresponds to the acquisition of determinate fate by the axillary meristem. The enrichment of MADS genes in the SM sample and their overrepresentation in cluster `r getClusterName('small', 'inc')` is consistent with their involvement in flower development in rice and other plants [`r I(pasteLabel("Figure ", figCount, "f_gsea"))`; *e.g.* @Kyozuka:2002vm; @Nagasawa:2003gu;  @Yamaguchi:2006br; @Yao:2008dy; @Dreni:2011bu; the role of MADS genes in Arabidopsis flower development was reviewed by @Prunet:2013ki]. Two recent studies in *A. thaliana* and the domesticated tomato, *Solanum lycopersicum*, used an LMD and RNAseq approach to investigate inflorescence development [@Park:2012dg; @Mantegazza:2014di]. These two datasets were used with the dataset presented in this article to compare the expression of the MIKC^C^-type class of MADS-box genes [@Henschel:2002ui], suggesting several differences between the species (`r I(pasteLabel("Figure S", s_figCount, "sf_madsTree"))`). *A. thaliana* *AGL6* is the only gene from the *AGL6*-like clade that is more highly expressed in indeterminate meristems than determinate meristems. *agl6* mutants have fewer axillary buds on the inflorescence stem, suggesting that *AGL6* promotes axillary meristem formation [@Huang:2012de]. The rice genes from the *AGL6*-like clade, *MFO* (*MADS6*) and *MADS17*, redundantly promote floral meristem determinancy [@Ohmori:2009kj], and are correspondingly enriched in determinate meristems. Expression of the putative tomato *AGL6*-like gene *SOLYC01G093960.2* is also enriched in determinate meristems, suggesting that the role of *AGL6* in axillary meristem formation may be specific to the Arabidopsis lineage. In contrast, consistent with their functions in floral organ specification in rice and *A. thaliana*, genes in the *AGL2*-like clade are generally more highly expressed in determinate meristems.

@@

This report describes changes in gene expression during the early stages of rice panicle development. Based on expression patterns of uncharacterized genes, transcription factors and genes with functions potentially related to hormone signalling, a switch between apical and axillary meristems was evident, followed by gradual changes during transitions between PBM and SM. This could be explained by the difference between the fates of the rachis meristem, which aborts after PBM differentiation [@Ikeda:2004tk], and the axillary meristems, which transition to determinate SM. The latter process may be comparable to the gradual meristem maturation during tomato inflorescence development [@Park:2012dg], rather than to the continual production of determinate floral meristems from the Arabidopsis inflorescence meristem. These results are expected to lead to mechanistic studies of the regulatory networks that establish gene expression in these tissues, as well as functional studies of the individual components, both of which will be vital for understanding the genetic factors influencing rice yield via panicle morphology.

@@

Samples harvested from the main stem from four developmental stages of panicle development were embedded in Paraplast X-TRA (Sigma-Aldrich) as described by @Huijser:1992wz. Digoxigenin-labelled antisense and sense RNA probes were generated with the DIG RNA Labeling Kit SP6/T7 (Roche) according to the manufacturer's instructions. To generate the probes, cDNA was amplified using the following primers: *LOC_Os01g04670*, 5′-GTGTCAAGGCATCGCCAAC-3′ and 5′-CATCAGCTGGCTGCTTTACC-3′; *LOC_Os10g04270*, 5′-TCCCAGTTGACCGAGAACTG-3′ and 5′-TGAACTTCCGTCACGAACTCC-3′; *LOC_Os10g05990*, 5′-CCTCCGGCAAAGAACTGATG-3′ and 5′-CCGTCATTGGGACTAGTTTGTCAG-3′. For detection of *LOC_Os09g27730*, a Digoxigenin-labelled LNA probe with the sequence 5′-TCTG{A}CGACG{T}GCG{A}C{T}GGT-3′ was synthesised (Eurogentec, Liège, Belgium). Hybridization was performed as described by @Coen:1990ux using NBT/BCIP (Roche) or VECTOR Blue Alkaline Phosphatase Substrate Kit (Vector Laboratories) for detection.

@@

Analysis of the sequenced libraries was carried out with `bash` and `R` scripts [@RALanguageandEn:2015wf], which were arranged into a pipeline using the `ruffus` package for `python3` [@Goodstadt:2010jp]. All of the code used for the analysis and to generate text, figures and tables for this report, along with the versions and paramaters of the software used, are available with revision history in a public GitHub repository at https://github.com/evoreprice/lmdPaper.

Briefly, adaptor trimming was performed with `cutadapt` [@Martin:2011eu] and reads were mapped against the MSUv7 genome and annotation [@Kawahara:2013gh; @Ouyang:2007il] downloaded from Phytozome 10.3 [@Goodstein:2012ho] using `STAR` [@Dobin:2012fg] in 2-pass mode. The `'quantMode'` argument of `STAR` was used for read counting. rRNA and tRNA contamination was estimated using `htseq-count` [@Anders:2014gf] to count the number of reads that mapped to regions annotated as tRNA or rRNA on the Rap-DB [@Ohyanagi:2006go] or to regions where reads simulated with `wgsim` (https://github.com/lh3/wgsim) from rRNA sequences in the TIGR Plant Repeats Database [@Ouyang:2004ia] also mapped. Differential expression analysis to calculate Log~2~-fold change (L~2~FC) values and transformed read counts was performed with the `DESeq2` package [@Love:2014do]. To determine a strict cut-off for unexpressed genes, expression values in transcripts per million were calculated for each gene [@Li:2010cl; @Wagner:2012jf]. These values were compared to pseudo-expression values calculated from intergenic regions of the genome, using intervals of a similar size distribution to those used for calculating the 'genic' expression values. For each library, the cut-off was placed at the 95^th^ percentile of the distribution of intergenic expression values. Transformed counts and L~2~FC values from expressed genes were used for soft clustering with the `Mfuzz` package [@Kumar:2007uw] and geneset enrichment analysis with the `gage` package [@Luo:2009kp] respectively. Gene lists were analysed using annotations from several databases, including TIGR [@Kawahara:2013gh], Oryzabase [@Kurata:2006gb] and OGRO [@Yamamoto:2012kw]. All plots were produced with the `ggplot2` package [@Wickham:2009tn].

@@

We thank James Tregear for feedback and discussion, and Jean-Luc Verdeil and Geneviève Conejero for assistance with laser microdissection at the PHIV imaging platform (CIRAD, Monpellier, France). This project was jointly supported by Fondation Agropolis (ANR-10-LABX-0001-01) and Fondazione Cariplo (project EVOREPRICE 1201-004).

>>>>

---
title: Gene expression profiling of reproductive meristem types in early rice inflorescences by laser microdissection
csl: csl/the-plant-journal.csl
header-includes:
  \usepackage{graphicx}
  \usepackage[utf8]{inputenc}
  \DeclareUnicodeCharacter{2265}{$\geq$}
output:
  word_document:
    fig_caption: yes
    reference_docx: csl/wordStyles.docx
  BiocStyle::html_document:
    dev: CairoPDF
    fig_caption: yes
    force_captions: yes
    highlight: pygments
    number_sections: yes
bibliography: bib/papersLib.bibtex
---

@@

In rice, the start of the reproductive phase and subsequent initiation of inflorescence development involves fast morphological transformations [described by @Ikeda:2004tk]. A detailed histological analysis of rice inflorescences was performed to select four morphologically distinct inflorescence meristem types for laser microdissection (LMD). After differentiation of the flag leaf, the first stage of reproductive development is the conversion of the shoot apical meristem (SAM) to the rachis meristem (RM). The first bract primordium was produced opposite the flag leaf (`r I(pasteLabel("Figure ", figCount, "f_lmdFigure"))`a). After the establishment of the RM, some cells differentiate into primary branch meristem (PBM) in the axils of newly developed bracts (`r I(pasteLabel("Figure ", figCount, "f_lmdFigure"))`b). After the formation of primary branches, bract growth ceased and primary branches elongated (ePBM) (`r I(pasteLabel("Figure ", figCount, "f_lmdFigure"))`c). During primary branch elongation, the PBM can give rise to axillary meristems (AM), which may differentiate into secondary and higher-order branches or be converted directly to spikelet meristem (SM). For this analysis, SM differentiation was considered as the final stage of panicle development (`r I(pasteLabel("Figure ", figCount, "f_lmdFigure"))`d). The PBM and secondary branch meristem (SBM) are both converted to a terminal SM. Each terminal or axillary SM produces one floret meristem (FM), which differentiates into a single floret. The RM, PBM and ePBM/AM stages are the indeterminate stages in which meristematic cells are maintained, whilst the SM has a determinate fate in which the stem cell activity will be lost and from which florets will differentiate [@Ikeda:2004tk].

@@

Several previous transcriptomic studies have measured gene expression in whole rice inflorescences at later stages of development than those described here [*e.g.* @Wang:2010im; @Sato:2011eh; @Sharma:2012kq; @Khanday:2013bz; @Jiang:2014im]. Another study used microarray experiments at earlier stages to highlight the importance of transcription factors in panicle development [@Furutani:2006hx]. LMD has also been used for precise control of developmental stage during the collection of whole panicle sections for microarrays [@Kobayashi:2012fi]. This identified three MADS genes, which are members of the *SQUA*-like (*FUL*-like) clade, that are co-expressed with *PANICLE PHYTOMER2* (*PAP2*/*MADS34*), and a quadruple knockdown of these four genes resulted in defects in inflorescence development [@Kobayashi:2012fi]. All four genes were detected in all stages in the RNAseq dataset (Data S1), confirming their expression in reproductive meristems. In contrast to previous transcriptomic studies, only the meristematic regions of young panicles were collected by LMD for the RNAseq dataset presented in this article, to restrict the measurement of gene expression to those tissues.

@@

Clusters `r getClusterName('dec', 'small')` and `r getClusterName('inc', 'small')` contain genes that change in expression between apical (RM) and axillary meristems (PBM to SM). Axillary meristem initiation requires auxin synthesis and transport in *Arabidopsis thaliana* and maize [reviewed by @Gallavotti:2013ki], and excess auxin may be inactivated by conjugation to amino acids by GH3 enzymes [@Staswick:2005cv], but these processes are not as well understood in rice. Cluster `r getClusterName('inc', 'small')` contains the auxin response factor *ARF6A* and the small auxin-up RNA *SAUR33*, as well as two ABC transporters (*ALS1* and *MDR12*), while the complementary cluster `r getClusterName('dec', 'small')` contains four uncharacterized genes with annotations relating to auxin (*IAA10* and *IAA16*, two auxin-responsive transcriptional regulators; *ABCB27*, an ABC transporter; and *LOC_Os01g63770*, an AUX/LAX transporter), and the GH3 gene related to auxin homeostasis, *GH3-8*. Transgenic rice plants presumably expressing *GH3-8* in the endogenous pattern of *FLO-LFY HOMOLOG OF RICE* (*RFL*) [*i.e.* PBM and SBM; @IkedaKawakatsu:2012co] have smaller panicles with fewer branches [@Yadav:2011db], and GH3-8 conjugates amino acids to auxin *in vitro* [@Ding:2008hv], suggesting that the panicle phenotype may be related to reduced auxin signalling. Although auxin response factors and Aux/IAA response genes are generally more strongly expressed at earlier stages (`r I(pasteLabel("Figure ", figCount, "f_gsea"))`), the presence of auxin-related genes in complementary clusters suggests that the transcriptional effects of auxin signalling are dependent on meristem type. Understanding of the relationship between auxin and meristem specification will require functional analysis of the components of the regulatory network involved in the response in these tissues.

Two *STMADS11*-like (*SVP*-like) MADS genes, *MADS22* and *MADS55*, decrease in expression between PBM and SM, whilst the third (*MADS47*) was not detected (`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`a). *MADS22* and *MADS55* are activated by the *G1*-Like (G1L/ALOG) transcriptional activator *G1L5* (*TAWAWA1*/*TAW1*), but activation of *MADS47* has not been reported [@Yoshida:2013ff]. Overexpression of *TAW1* increases the production of secondary branches and leads to the formation of tertiary branches, resulting in higher grain yield [@Yoshida:2013ff]. *TAW1* was recovered from the same cluster as *MADS55*, and the presence and co-expression of genes that are involved in inflorescence development along with other TF genes indicates that the clusters capture biologically meaningful groups of genes, including uncharacterized signalling components. Only three ALOG genes (which are annotated with the Pfam domain PF04852) were detected in the dataset. Two were retrieved from clusters `r getClusterName('dec', 'small')` and `r getClusterName('dec', 'dec')` (*G1L2* and *TAW1* respectively), and the third (*G1L1*) has a similar expression pattern (`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`b). *taw1* missense mutants have small inflorescences with reduced grain yield, but this phenotype was stronger when *TAW1* was targeted by RNA interference [@Yoshida:2013ff]. This may suggest functional redundancy between the three co-expressed ALOG genes, *G1L1*, *G1L2* and *TAW1*, with more than one affected by the RNAi construct used against *TAW1*. Although the increase in secondary branches linked to *TAW1* overexpression suggests that it acts by suppression of SM identity, perhaps via downstream factors that may promote BM activity such as the *STMADS11*-like genes, *taw1* plants also have fewer primary branches [@Yoshida:2013ff], and the peak expression in the RM of *G1L1*, *G1L2*, *TAW1*, *MADS22* and *MADS55* suggests an additional role before the transition from apical to axillary meristem. The difference in expression of annotated transcription factor genes between RM and PBM samples and the recovery of co-expression clusters of primarily uncharacterized genes that appear to be either switched on or off between the RM and the PBM indicate a significant change in gene expression between apical and axillary meristems.

@@

Compared to the switch from apical to axillary meristem, more gradual changes in gene expression were apparent during the sequential transition from PBM to SM (`r I(pasteLabel("Figure ", figCount, "f_mfuzzClusters"))`). This transition corresponds to the acquisition of determinate fate by the axillary meristem. The enrichment of MADS genes in the SM sample and their overrepresentation in cluster `r getClusterName('small', 'inc')` is consistent with their involvement in flower development in rice and other plants [`r I(pasteLabel("Figure ", figCount, "f_gsea"))`; *e.g.* @Kyozuka:2002vm; @Nagasawa:2003gu;  @Yamaguchi:2006br; @Yao:2008dy; @Dreni:2011bu; the role of MADS genes in Arabidopsis flower development was reviewed by @Prunet:2013ki]. Two recent studies in *A. thaliana* and the domesticated tomato, *Solanum lycopersicum*, used an LMD and RNAseq approach to investigate inflorescence development [@Park:2012dg; @Mantegazza:2014di]. These two datasets were used with the dataset presented in this article to compare the expression of the MIKC^C^-type class of MADS-box genes [@Henschel:2002ui], suggesting several differences between the species (`r I(pasteLabel("Figure S", s_figCount, "sf_madsTree"))`). *A. thaliana* *AGL6* is the only gene from the *AGL6*-like clade that is more highly expressed in indeterminate meristems than determinate meristems. *agl6* mutants have fewer axillary buds on the inflorescence stem, suggesting that *AGL6* promotes axillary meristem formation [@Huang:2012de]. The rice genes from the *AGL6*-like clade, *MFO* (*MADS6*) and *MADS17*, redundantly promote floral meristem determinancy [@Ohmori:2009kj], and are correspondingly enriched in determinate meristems. Expression of the putative tomato *AGL6*-like gene *SOLYC01G093960.2* is also enriched in determinate meristems, suggesting that the role of *AGL6* in axillary meristem formation may be specific to the Arabidopsis lineage. In contrast, consistent with their functions in floral organ specification in rice and *A. thaliana*, genes in the *AGL2*-like clade are generally more highly expressed in determinate meristems. *PAP2*, which was strongly expressed at all stages with higher expression in the RM and PBM (Data S1), acts with another *AGAMOUS-LIKE2*-like (*AGL2*-like/*SEPELLATA*) MADS gene, *LEAFY HULL STERILE 1* (*LHS1*/*MADS1*), to specify floral organs [@Gao:2010iz]. *pap2* mutants also have altered primary and secondary branch number, suggesting an earlier role is spikelet meristem specification [@Gao:2010iz; @Kobayashi:2010hn]. *LHS1*, which is involved in floral meristem determination and floral organ development [@Jeon:2000tj; @Agrawal:2005ho], was enriched in the SM (`r I(pasteLabel("Figure S", s_figCount, "sf_madsTree"))`).

@@

The molecular mechanisms controlling the meristematic activities that guide panicle development in rice remain largely uncharacterized. The dataset presented here addresses this by describing both the transcriptome profiles associated with specific meristem identities and the changes in gene expression that occur during the early stages of panicle development. A marked expression switch was evident between apical and axillary meristems, followed by gradual changes during transitions between PBM and SM. This could be explained by the difference between the fates of the rachis meristem, which aborts after PBM differentiation [@Ikeda:2004tk], and the axillary meristems, which undergo transition to determinate SMs. The latter process may be comparable to the gradual meristem maturation observed during tomato inflorescence development [@Park:2012dg], rather than to the continual production of determinate floral meristems characteristic of the Arabidopsis inflorescence meristem. Promising uncharacterized genes were identified, notable examples including transcription factor genes with expression patterns similar to known regulators of panicle architecture and putative hormone-related such as the auxin-responsive genes. Functional studies of these genes will reveal the mechanics of the regulatory networks that establish gene expression patterns in meristems and their control by hormone signalling. Genes with expression specific to certain meristem types were also identified, which may prove valuable as marker genes, for example in developmental studies of mutants affected in determination of reproductive meristem identity. Combined with current technology for the creation of mutants, this dataset promises to accelerate research on the molecular mechanisms controlling panicle development, a vital component of efforts to achieve the sustainable increase in grain yield required to address population growth.

@@

Samples harvested from the main stem from four developmental stages of panicle development were embedded in Paraplast X-TRA (Sigma-Aldrich) as described by @Huijser:1992wz. digoxigenin-labelled antisense and sense RNA probes were generated with the DIG RNA Labeling Kit SP6/T7 (Roche) according to the manufacturer's instructions. To generate the probes, cDNA was amplified using the following primers: *LOC_Os01g04670*, 5′-GTGTCAAGGCATCGCCAAC-3′ and 5′-CATCAGCTGGCTGCTTTACC-3′; *LOC_Os10g04270*, 5′-TCCCAGTTGACCGAGAACTG-3′ and 5′-TGAACTTCCGTCACGAACTCC-3′; *LOC_Os10g05990*, 5′-CCTCCGGCAAAGAACTGATG-3′ and 5′-CCGTCATTGGGACTAGTTTGTCAG-3′. For detection of *LOC_Os09g27730*, a Digoxigenin-labelled LNA probe with the sequence 5′-TCTG{A}CGACG{T}GCG{A}C{T}GGT-3′ was synthesised (Eurogentec, Liège, Belgium). Hybridization was performed as described by @Coen:1990ux using NBT/BCIP (Roche) or VECTOR Blue Alkaline Phosphatase Substrate Kit (Vector Laboratories) for detection.

@@

Analysis of the sequenced libraries was carried out with `bash` and `R` scripts [@RALanguageandEn:2015wf], which were arranged into a pipeline using the `ruffus` package for `python3` [@Goodstadt:2010jp]. All of the code used for the analysis and to generate text, figures and tables for this report, along with the versions and parameters of the software used, are available with revision history in a public GitHub repository at [https://github.com/evoreprice/lmdPaper](https://github.com/evoreprice/lmdPaper).

Briefly, adaptor trimming was performed with `cutadapt` [@Martin:2011eu] and reads were mapped against the MSUv7 genome and annotation [@Kawahara:2013gh; @Ouyang:2007il] downloaded from Phytozome 10.3 [@Goodstein:2012ho] using `STAR` [@Dobin:2012fg] in 2-pass mode. The `'quantMode'` argument of `STAR` was used for read counting. rRNA and tRNA contamination was estimated using `htseq-count` [@Anders:2014gf] to count the number of reads that mapped to regions annotated as tRNA or rRNA on the Rap-DB [@Ohyanagi:2006go] or to regions where reads simulated with `wgsim` ([https://github.com/lh3/wgsim](https://github.com/lh3/wgsim)) from rRNA sequences in the TIGR Plant Repeats Database [@Ouyang:2004ia] also mapped. Differential expression analysis to calculate Log~2~-fold change (L~2~FC) values and transformed read counts was performed with the `DESeq2` package [@Love:2014do]. To determine a strict cut-off for unexpressed genes, expression values in transcripts per million were calculated for each gene [@Li:2010cl; @Wagner:2012jf]. These values were compared to pseudo-expression values calculated from intergenic regions of the genome, using intervals of a similar size distribution to those used for calculating the 'genic' expression values. For each library, the cut-off was placed at the 95^th^ percentile of the distribution of intergenic expression values. Transformed counts and L~2~FC values from expressed genes were used for soft clustering with the `Mfuzz` package [@Kumar:2007uw] and geneset enrichment analysis with the `gage` package [@Luo:2009kp] respectively. Gene lists were analysed using annotations from several databases, including TIGR [@Kawahara:2013gh], Oryzabase [@Kurata:2006gb] and OGRO [@Yamamoto:2012kw]. All plots were produced with the `ggplot2` package [@Wickham:2009tn].

@@

We thank James Tregear for feedback and discussion, and Jean-Luc Verdeil and Geneviève Conejero for assistance with laser microdissection at the PHIV imaging platform (CIRAD, Monpellier, France). This project was jointly supported by Fondation Agropolis through the *Investissements d'avenir* programme (ANR-10-LABX-0001-01) and Fondazione Cariplo under the EVOREPRICE 1201-004 project.
